---
output: html_document
---
Assignment no 1 by Mait22  
==========================  


##Loading in the data
Following r code loads in the data set "activity.csv" placed in working directory
of Your machine.

```{r}
work_dir <- getwd()
file_path <- paste(work_dir,"/activity.csv",sep = "")
activity <- read.csv(file_path,header = TRUE,na.strings = NA,
                     stringsAsFactors = FALSE)
```


##Calculating the total number of steps taken per day 
I use plyer package to complete this task - install it if needed.

```{r}
#install.packages("plyer") ##uncomment and install the package if needed
library(plyr)
descriptives <- ddply(activity,~date,summarise,tot_steps=sum(steps))

steps_per_day_median <- median(descriptives$tot_steps,na.rm = TRUE)
steps_per_day_mean <- as.integer(round(mean(descriptives$tot_steps,na.rm = TRUE),1))

hist(as.numeric(descriptives$tot_steps),ylab = "Frequency",xlab= "Total steps in given day",main = "Histogram of steps taken per day in total",breaks = 20)


```

The median number of total steps taken per day is `r steps_per_day_median`  
The mean number of total steps taken per day is `r steps_per_day_mean`


#Plotting the average daily activity by time interval

Following r code calculates the daily average activity by interval  
```{r}
descriptives_by_interval <-ddply(activity,~interval,summarise,mean=mean(steps,na.rm = TRUE),
                                                  sd=sd(steps,na.rm = TRUE))

descriptives_by_interval <- arrange(descriptives_by_interval, desc(mean))

```

Following table presents the top 5 intervals with highest level of average steps taken.
```{r}
library(xtable) #Install if needed
library("knitr") #Install if needed
library("xtable") #Install if needed
xt <- xtable(descriptives_by_interval[1:5,])
kable(xt, caption = "Top 5 intervals with highest degree of steps",digits = 2,align = "l")
```

Plotting steps taken by interval across all days in dataset
```{r}

descriptives_by_interval <- arrange(descriptives_by_interval, desc(interval))

plot(descriptives_by_interval$interval,descriptives_by_interval$mean, type = "l",xlab = "Interval",ylab = "Steps taken", main = "Steps taken by interval",lwd = 2, col ="green")
```



```{r}
num_of_na <- sum(is.na(activity[,1]))
```
#Imputing the missing data

In total the data had `r num_of_na` rows with missing activity data. 

####Following strategy was used to impute for missing data:  
-  Averages for each time intervals were calculated  
-  For loop was constructed (see following r code chunk) to check if current row has missing value, if so the average value of given row's interval was used to fill in for missing data 

```{r}

missing_values_imputed <- activity

for (i in 1:length(missing_values_imputed[,1])){
       if (is.na(missing_values_imputed[i,1]) == TRUE)
              {
              missing_values_imputed[i,1] <- descriptives_by_interval[descriptives_by_interval[,1] == missing_values_imputed[i,3],2]
       }
}

```

```{r}

descriptives_new <- ddply(missing_values_imputed,~date,summarise,tot_steps=sum(steps))

new_median <- as.integer(median(descriptives_new$tot_steps,na.rm = TRUE))
new_mean <- as.integer(round(mean(descriptives_new$tot_steps,na.rm = TRUE),1))

```


After the imputation of data the median value of steps per day was `r new_median` and new average of steps per day was `r new_mean`.  
Histogram of activity after imputing for missing data is presented next:

```{r}


hist(as.numeric(descriptives_new$tot_steps),ylab = "Frequency",xlab= "Total steps in given day",main = "Histogram of steps taken per day in total",breaks = 20)

```


##Making plots to decide if there is difference in activity patterns between weekdays and weekends


```{r,fig.height=12}
#Making the plot
missing_values_imputed$date <- weekdays(as.Date(missing_values_imputed$date))
weekdays <- missing_values_imputed[missing_values_imputed$date %in% c("Monday","Tuesday","Wednesday","Thursday","Friday"),]
weekend <-  missing_values_imputed[missing_values_imputed$date %in% c("Saturday","Sunday"),]      

descriptives_weekdays <- descriptives <- ddply(weekdays,~interval,summarise,mean=mean(steps,na.rm = TRUE),
                                                  sd=sd(steps,na.rm = TRUE))

descriptives_weekends <- descriptives <- ddply(weekend,~interval,summarise,mean=mean(steps,na.rm = TRUE),
                                                  sd=sd(steps,na.rm = TRUE))

descriptives_weekdays <- arrange(descriptives_weekdays, interval)
descriptives_weekends <- arrange(descriptives_weekends, interval)

par(mfrow =c(2,1))
plot(descriptives_weekdays$interval,descriptives_weekdays$mean, type = "l",xlab = "Interval",ylab = "Steps taken", main = "Weekday",lwd = 2, col ="green")
plot(descriptives_weekends$interval,descriptives_weekends$mean, type = "l",xlab = "Interval",ylab = "Steps taken", main = "Weekend",lwd = 2, col ="green")
```


From the visual inspection of graphs it seems that in weekends the later time intervals have markedly higher activity level compared to in-week days.